The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-08-29 manuscript.Rmd
add latitudinal gradient figure

<!-- need to add latitudinal gradient map -->
On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig. \@ref(fig:map-diversity-sampling)B).

@@

@@

Spatial patterns of species richness are strongly confounded by sampling effort. There is a strong correlation between the number of samples and number of species per cell, and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)).

Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change. Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions. To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008]. Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS. Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.

Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Explorations in the late ninetieth century led to an exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. <!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates. For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].

@@

Further biases exist across phyla and body size. Large-bodied and easily acccessible (e.g. dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna. Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records. This underrepresentation is manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for the `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`). Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012]. Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.

The proportion of samples that are identified to species show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig. \@ref(fig:identification-through-time)). There is a marked break in the central trend arond the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections. For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.

@@

>>>>

Spatial distribution of species richness with latitude differ markedly between the Pacific and the Atlantic coasts. On the Pacific coast, richness is lowest around 40° of latitude, while it is where it is the highest on the Atlantic coast (Fig. \@ref(fig:map-diversity-sampling)B, Fig. \@ref(fig:latitudinal-gradient)).

@@

```{r latitudinal-gradient, fig.cap='Latitudinal gradient in species richness for the Pacific and Atlantic oceans (Gulf of Mexico records excluded). Trend lines are B-splines of the 3rd degree polynomial weighted by the square-root of the number of records. (\\#fig:latitudinal-gradient)'}
knitr::include_graphics("figures/combined_gradient.pdf")
```

@@

Spatial patterns of species richness are however strongly confounded by sampling effort. There is a high correlation between the number of samples and number of species per cell used in Fig. \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)).

Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change. Predictive ecological models forecasting changes in distributions are trained on environmental conditions encountered by species in their current distributions. To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008]. Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS. Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.

Another indication of the inadequacy of the sampling is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Explorations in the late ninetieth century led to an exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.

@@

Further biases exist across phyla and body size. Large-bodied and easily acccessible (e.g. dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna. Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records. This underrepresentation is also manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for these `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` others. Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012]. Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.

The proportion of samples that are identified to species in museum collections show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig. \@ref(fig:identification-through-time)). There is a marked break in the central trend arond the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections. For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.

@@

<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates. For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].
